---
title: "1.transform (exercises)"
author: "Joao Lopes"
date: "2026-03-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, eval=FALSE)

```

\newpage

\tableofcontents

***
\newpage

# 1. NUMERIC VECTORS

[from https://r4ds.hadley.nz/numbers.html]

[see https://jrnold.github.io/r4ds-exercise-solutions/transform.html]

[see https://mine-cetinkaya-rundel.github.io/r4ds-solutions/numbers.html]

```{r 1_numerical, eval=TRUE, warning=FALSE, message=FALSE}
library("nycflights13")
library("tidyverse")

```


## 1.1. COUNTS

a\) How can you use count() to count the number of rows with a missing value for a given variable?
```{r 1_1_1_numerical}
#number of missing values for a given variable
flights |> mutate(is_miss = is.na(dep_time)) |> count(is_miss)

#number of missing values for at least one variable
flights |> mutate(is_miss = if_any(everything(), ~ is.na(.))) |> count(is_miss)

#number of missing values for each variable
flights |>
  select(where(~ any(is.na(.)))) |>
  summarise(across(everything(), ~ sum(is.na(.))))
  
```

b\) Expand the following calls to count() to instead use group_by(), summarize(), and arrange():

1\. flights |> count(dest, sort = TRUE)
```{r 1_1_2_numerical}
flights |> 
  group_by(dest) |>
  summarize(n = n()) |>
  arrange(desc(n))

```

2\. flights |> count(tailnum, wt = distance)
```{r 1_1_3_numerical}
flights |> 
  group_by(tailnum) |> 
  summarize(miles = sum(distance))

```

## 1.2. NUMERIC TRANSFORMATION

a\) Explain in words what each line of the following code does:
```{r 1_2_1_numerical, echo=TRUE, eval=FALSE}
flights |> 
  group_by(hour = sched_dep_time %/% 100) |> 
  summarize(prop_cancelled = mean(is.na(dep_time)), n = n()) |> 
  filter(hour > 1) |> 
  ggplot(aes(x = hour, y = prop_cancelled)) +
  geom_line(color = "grey50") + 
  geom_point(aes(size = n))

```
```{r 1_2_2_numerical}
#use dataset flights
flights |>
  #group flights per hour of schedule departure
  group_by(hour = sched_dep_time %/% 100) |>
  summarize(
    #calculate proportion of NAs in dep_time
    prop_cancelled = mean(is.na(dep_time)),
    #calculate total number of flights
    n = n()) |>
  #filter-out flights schedule for before 1:00
  filter(hour > 1) |>
  #select variables hour and prop_cancelled
  ggplot(aes(x = hour, y = prop_cancelled)) +
  #create lineplot of variables
  geom_line(color = "grey50") +
  #create scatterpoints of variables with size = n
  geom_point(aes(size = n))

```

b\) Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. You can see the basic problem by running the code below: there’s a gap between each hour.
```{r 1_2_3_numerical, echo=TRUE, eval=TRUE, fig.align='center'}
flights |> 
  filter(month == 1, day == 1, !is.na(dep_delay)) |> 
  ggplot(aes(x = sched_dep_time, y = dep_delay)) +
  geom_point()

```
Convert them to a more truthful representation of time (either fractional hours or minutes since midnight).
```{r 1_2_4_numerical}
flights |>
  mutate(
    hour = sched_dep_time %/% 100,  #extract hours of schedule departure
    minute = sched_dep_time %% 100, #extract minutes of schedule departure
#   time1 = hour*60 + minute        #minutes since midnight
    time1 = hour + minute/60        #fractional hours since midnight
  ) |>
  filter(month == 1, day == 1, !is.na(dep_delay)) |>
  ggplot(aes(x = time1, y = dep_delay)) +
  geom_point() + 
# labs(x = "minutes since midnight")
  labs(x = "fractional hours since midnight")
```

c\) Round dep_time and arr_time to the nearest five minutes
```{r 1_2_5_numerical}
flights |>
  mutate(
    #extract hours of departure
    dep_time_hh = dep_time %/% 100,
    #extract minutes of departure in 5min interval
    dep_time_mm = round(dep_time %% 100 / 5) * 5,
    #compute departure time using 5min interval
    dep_time2 = if_else(
      dep_time_mm <= 55,
      dep_time_hh * 100 + dep_time_mm, #rounded minutes less than 60
      (dep_time_hh + 1) * 100          #rounded minutes equal to 60
    ),
    .keep = "used"
  )

```

## 1.3. GENERAL TRANSFORMATION

a\) Find the 12 most delayed flights using a ranking function. How do you want
to handle ties? Carefully read the documentation for min_rank().
```{r 1_3_1_numerical, eval=FALSE}
?min_rank
?base::rank

```
```{r 1_3_2_numerical}
flights |>
  mutate(ranks = min_rank(desc(dep_delay))) |>
  select(tailnum, dep_delay, ranks) |>
  arrange(ranks) |>
  filter(ranks <= 12)

#using row_number()
flights |>
  mutate(ranks = row_number(desc(dep_delay))) |>
  select(tailnum, dep_delay, ranks) |>
  arrange(ranks) |>
  filter(ranks <= 12)

```

b\) Which plane (tailnum) has the worst on-time record?
```{r 1_3_3_numerical}
#only select the worst on-time record overall
flights |>
  filter(!is.na(dep_delay)) |>           #filter out NA values
  filter(dep_delay == max(dep_delay)) |> #filter in maximum value
  select(tailnum, dep_delay)             #select variables tailnum and dep_delay

#calculate worst on-time record for each plane
flights |>
  filter(!is.na(dep_delay)) |>     #filter out NA values
  group_by(tailnum) |>             #group by "tailnum"
  summarize(
    dep_delay_rec = max(dep_delay) #calculate maximum delay per group
  ) |>
  arrange(desc(dep_delay_rec))

```

c\) At which hour should you fly if you want to avoid more than one hour delays as much as possible?
```{r 1_3_4_numerical}
flights |>
  filter(!is.na(dep_delay)) |>
  group_by(hour) |>
  summarise(
    n_delay = sum(dep_delay > 60),             #n_delay; number of delays
    p_delay = mean(dep_delay > 60),            #p_delay: proportion of delays
    m_delay = mean(dep_delay[dep_delay > 0]),  #m_delay; mean delay
    t_delay = sum(dep_delay[dep_delay > 0])    #t_delay: total minutes delayed
  )
  
```

d\) What does flights |> group_by(dest) |> filter(row_number() < 4) do? What does flights |> group_by(dest) |> filter(row_number(dep_delay) < 4) do?
```{r 1_3_5_numerical, eval=FALSE}
?row_number

```
```{r 1_3_6_numerical}
#flights |> group_by(dest) |> filter(row_number() < 4)
flights |>
  #group by "dest"
  group_by(dest) |>
  #filter for the first 3 elements per group
  filter(row_number() < 4) |>
  select(dest, everything()) |> arrange(dest)

#flights |> group_by(dest) |> filter(row_number(dep_delay) < 4)
flights |>
  #group by "dest"
  group_by(dest) |>
  #filter for the 3 smallest departure delays per group
  filter(row_number(dep_delay) < 4) |>
  select(dest, everything()) |> arrange(dest)

```

e\) For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.
```{r 1_3_7_numerical}
#total minutes of delay for each destination
flights |>
  #filter out early departures
  filter(dep_delay > 0) |>
  #group by "dest"
  group_by(dest) |>
  #sum up delay time per group
  summarize(t_delay = sum(dep_delay, na.rm = TRUE))

#proportion of total delay per flight for each destination
flights |>
  #filter out early departures
  filter(dep_delay > 0) |>
  #group by "flight"
  group_by(dest, origin, carrier, flight) |>
  #sum up delay time per group
  summarize(t_delay = sum(dep_delay), .groups = "drop") |>
  #group by "dest"
  group_by(dest) |>
  #calculate proportion of delay per group
  mutate(p_delay = round(t_delay/sum(t_delay),3)) |>
  arrange(dest, desc(p_delay)) |>
  select(carrier, flight, origin, dest, p_delay)

```

f\) Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using lag(), explore how the average flight delay for an hour is related to the average delay for the previous hour.
```{r 1_3_8_numerical, echo=TRUE, eval=FALSE}
flights |> 
  mutate(hour = dep_time %/% 100) |> 
  group_by(year, month, day, hour) |> 
  summarize(
    dep_delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  filter(n > 5)
  
```
```{r 1_3_9_numerical}
flights_dep_lag <- flights |>
  #extract hour of departure
  mutate(hour = dep_time %/% 100) |>
  #group by "year", "month", "day" and "hour"
  group_by(year, month, day, hour) |>
  summarize(
    #calculate average delay per group
    dep_delay = mean(dep_delay, na.rm = TRUE),
    #calculate number of observations per group
    n = n(),
    .groups = "drop"
  ) |>
  #filter for groups with more than 5 observations
  filter(n > 5) |>
  #make sure data is order by time
  arrange(year, month, day, hour) |>
  #create variable of previous average delay
  mutate(dep_delay_lag = lag(dep_delay, n = 1)) |>
  #filter out no delays and no previous time point
  filter(dep_delay > 0, !is.na(dep_delay_lag))

#calculate correlation
flights_dep_lag |>
  summarize(cor(dep_delay, dep_delay_lag, use = "complete.obs"))

#create visualization
flights_dep_lag |>
  ggplot(aes(x = dep_delay_lag, y = dep_delay)) +
  geom_bin2d(bins = 50) +
  stat_smooth(method = "lm", formula = "y ~ x", se = FALSE)
    
rm(flights_dep_lag)

```

g\) Look at each destination. Can you find flights that are suspiciously fast (i.e. flights that represent a potential data entry error)? Compute the air time of a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?
```{r 1_3_10_numerical, fig.height=10, fig.width=10}
#visualize air time distributions for each "origin" and "dest"
flights |>
  ggplot(aes(x = air_time, y = dest)) +
  geom_boxplot() +
  facet_wrap(~ origin, ncol = 3) +
  theme(axis.text = element_text(size = 6))

```
```{r 1_3_11_numerical}
#extreme air time flights for each "origin" and "dest"
flights |>
  #group by "origin" and "dest"
  group_by(origin, dest) |>
  mutate(
    #calculate Q25(air_time) per group
    at_Q25 = quantile(air_time, prob = 0.25, na.rm = TRUE),
    #calculate Q50(air_time) per group
    at_Q50 = quantile(air_time, prob = 0.50, na.rm = TRUE),
    #calculate Q75(air_time) per group
    at_Q75 = quantile(air_time, prob = 0.75, na.rm = TRUE),
    #calculate suspiciousness
    at_flg = air_time < at_Q25 - 2 * (at_Q75 - at_Q25)
  ) |>
  filter(at_flg) |>
  select(origin, dest, tailnum, air_time, at_Q25, at_Q50, at_Q75) |>
  arrange(desc(air_time))

#most delayed flights for each "origin" and "dest"
flights |>
  #group by "origin" and "dest"
  group_by(origin, dest) |>
  mutate(
    #calculate minimum air time per group
    at_min = min(air_time, na.rm = TRUE),
    #calculate air_time relative to minimum per group
    at_rel = air_time/at_min,
    #calculate the top most delayed flight per group
    at_flg = min_rank(desc(at_rel)) == 1
  ) |>
  filter(at_flg) |>
  select(origin, dest, tailnum, air_time, at_min, at_rel) |>
  arrange(desc(at_rel))

```

h\) Find all destinations that are flown by at least two carriers. Use those destinations to come up with a relative ranking of the carriers based on their performance for the same destination.
```{r 1_3_12_numerical}
flights |>
  #filter out NA values 
  filter(!is.na(air_time)) |>
  #group by "dest"
  group_by(dest) |>
  #calculate number of observations per group
  mutate(n_carriers = n_distinct(carrier)) |>
  #filter for groups with at least 2 observations
  filter(n_carriers > 1) |>
  #group by "carrier"
  group_by(dest, origin, carrier) |>
  #calculate performance
  summarize(at_mean = mean(air_time)) |>
  group_by(dest, origin) |>
  #calculate rank of performance per group
  mutate(at_rnk = min_rank(at_mean)) |>
  select(dest, origin, carrier, at_mean, at_rnk) |>
  arrange(dest, origin, at_rnk)

```

***
\newpage

# 2. FACTORS

[from https://r4ds.hadley.nz/factors.html]

[see https://jrnold.github.io/r4ds-exercise-solutions/factors.html]

[see https://mine-cetinkaya-rundel.github.io/r4ds-solutions/factors.html]

```{r 2_factors, warning=FALSE, message=FALSE}
library("tidyverse")

```

## 2.1. BASICS
[no exercises]

## 2.2. DATASET gss_cat

a\) Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?
```{r 2_2_1_factors}
#default plot
gss_cat |>
  ggplot(aes(x = rincome)) +
  geom_bar()

#improved plot
gss_cat |>
  ggplot(aes(y = fct_relevel(rincome, "Not applicable"))) +
  geom_bar()

```

b\) What is the most common religion in this survey? What’s the most common political party?
```{r 2_2_2_factors}
#most common religion
gss_cat |> count(relig, sort = TRUE)

#most common political party
gss_cat |> count(partyid, sort = TRUE)
```

c\) Which religion does denomination (denom) apply to? How can you find out with a table? How can you find out with a visualization?
```{r 2_2_3_factors}
#check possible values of "denom"
gss_cat |> count(denom) |> print(n = Inf)

#check values of "relig" and "denom"
x <- c("No answer", "Other", "Don't know", "Not applicable", "No denomination")
gss_cat |> filter(!denom %in% x) |> count(relig, denom) |> print(n = Inf)

#visualization using geom_bar()
gss_cat |>
  ggplot(aes(y = relig, fill = denom)) +
  geom_bar() +
  theme(legend.position = "bottom", legend.box = "vertical")
  
#visualization using geom_count()
gss_cat |>
  ggplot(aes(x = relig, y = denom)) +
  geom_count() +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(axis.text.y = element_text(size = 7))

```

## 2.3. MODIFYING FACTOR ORDER

a\) For each factor in gss_cat identify whether the order of the levels is arbitrary or principled.
```{r 2_3_1_factors}
gss_cat |> count(marital) #somewhat principled
gss_cat |> count(race)    #principled by count of observations
gss_cat |> count(rincome) #principled
gss_cat |> count(partyid) #principled
gss_cat |> count(relig)   #arbitrary
gss_cat |> count(denom)   #arbitrary

```

b\) Why did moving “Not applicable” to the front of the levels move it to the bottom of the plot?
```{r 2_3_2_factors}
gss_cat |>
  #move closer to origin of axis
  ggplot(aes(y = fct_relevel(rincome, "Not applicable"))) +
  geom_bar()

```

## 2.4. MODIFYING FACTOR LEVELS

a\) How have the proportions of people identifying as Democrat, Republican, and Independent changed over time?
```{r 2_4_1_factors}
gss_cat_mod <- gss_cat |>
  mutate(
    partyid = fct_collapse(partyid,
      "other" = c("No answer", "Don't know", "Other party"),
      "rep" = c("Strong republican", "Not str republican"),
      "ind" = c("Ind,near rep", "Independent", "Ind,near dem"),
      "dem" = c("Not str democrat", "Strong democrat")
    )
  )

#visualization using geom_bar(position = "fill")
gss_cat_mod |>
  ggplot(aes(x = year, fill = partyid)) +
  geom_bar(position = "fill")

#visualization using geom_point() and geom_line()
gss_cat_mod |>
  count(year, partyid) |>
  group_by(year) |>
  mutate(p = n/sum(n)) |>
  ggplot(aes(x = year, y = p, colour = partyid)) +
  geom_point() +
  geom_line()

rm(gss_cat_mod)
```

b\) How could you collapse rincome into a small set of categories?
```{r 2_4_2_factors}
gss_cat |> count(rincome)
gss_cat |>
  mutate(
    rincome = fct_collapse(rincome,
      "Not applicable" = "Not applicable",
      "N/A"            = c("No answer",
                           "Don't know",
                           "Refused"),
      "$25000 or more" = "$25000 or more",
      "$10000 - 24999" = c("$20000 - 24999",
                           "$15000 - 19999",
                           "$10000 - 14999"),
      "$5000 - 9999"   = c("$8000 to 9999",
                           "$7000 to 7999",
                           "$6000 to 6999",
                           "$5000 to 5999"),
      "Lt $5000"       = c("$4000 to 4999",
                           "$3000 to 3999",
                           "$1000 to 2999",
                           "Lt $1000")
    )
  ) |>
  count(rincome)

```

c\) Notice there are 9 groups (excluding other) in the fct_lump example above. Why not 10? (Hint: type ?fct_lump, and find the default for the argument other_level is “Other”.)
```{r 2_4_3_factors, eval=FALSE}
?fct_lump

```
```{r 2_4_4_factors}
#keep other_level="Other"
gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 10)) |>
  count(relig) |>
  print(n = Inf)

#change to other_level="Other answer"
gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 10, other_level="Other answer")) |>
  count(relig) |>
  print(n = Inf)

```

***
\newpage
  
# 3. LOGICAL VECTORS

[from https://r4ds.hadley.nz/logicals.html]

[see https://jrnold.github.io/r4ds-exercise-solutions/vectors.html]

[see https://mine-cetinkaya-rundel.github.io/r4ds-solutions/logicals.html]

```{r 3_logical, warning=FALSE, message=FALSE}
library("nycflights13")
library("tidyverse")

```

## 3.1. COMPARISONS

a) How does dplyr::near() work? Type near to see the source code. Is sqrt(2)^2 near 2?
```{r 3_1_1_logical, eval=FALSE}
?near
near
```
```{r 3_1_2_logical}
near(sqrt(2)^2, 2)
near(sqrt(2)^2, 2, tol = 1e-16)

```

b) Use mutate(), is.na(), and count() together to describe how the missing values in dep_time, sched_dep_time and dep_delay are connected.?
```{r 3_1_3_logical}
flights |>
  mutate(
    #calculate number of missing values for dep_time
    dt_na = is.na(dep_time),
    #calculate number of missing values for sched_dep_time
    sdt_na = is.na(sched_dep_time),
    #calculate number of missing values for dep_delay
    dd_na = is.na(dep_delay),
    .keep="used"
  ) |>
  #group by "dt_na", "sdt_na" and "dd_na"
  group_by(dt_na, sdt_na, dd_na) |>
  count()

```

## 3.2. BOOLEAN ALGEBRA

a\) Find all flights where arr_delay is missing but dep_delay is not. Find all flights where neither arr_time nor sched_arr_time are missing, but arr_delay is.
```{r 3_2_1_logical}
#arr_delay is missing but dep_delay is not
flights |>
  filter(is.na(arr_delay) & !is.na(dep_delay))

#neither arr_time nor sched_arr_time are missing, but arr_delay is
flights |>
  filter(!is.na(arr_time) & !is.na(sched_arr_time) & is.na(arr_delay))

```

b\) How many flights have a missing dep_time? What other variables are missing in these rows? What might these rows represent?
```{r 3_2_2_logical}
#number of flights with missing dep_time
flights |> filter(is.na(dep_time)) |> count()

#look at flights with missing dep_time
flights |>
  #filter in cancelled flights
  filter(is.na(dep_time)) |>
  #select any columns with missing values
  select(where(~ any(is.na(.)))) |>
  #calculate number of missing values
  summarise(across(everything(),~ sum(is.na(.))))

```

c\) Assuming that a missing dep_time implies that a flight is cancelled, look at the number of cancelled flights per day. Is there a pattern? Is there a  connection between the proportion of cancelled flights and the average delay of non-cancelled flights?
```{r 3_2_3_logical}
#search for pattern in number of cancelled flights per day
flights |>
  #group by "year", "month", "day"
  group_by(year, month, day) |>
  summarise(
    #number of missing in "dep_time" per groups
    n_cancel = sum(is.na(dep_time)),
    .groups = "drop"
  ) |> 
  mutate(
    #calculate day of the year
    year_day = row_number()
  ) |>
  ggplot(aes(x = year_day, y = n_cancel)) +
  geom_line() +
  geom_point()

#search for connection between proportion of cancelled and average delay
flights |>
  #group by "year", "month", "day"
  group_by(year, month, day) |>
  summarise(
    #proportion of missing in "dep_time" per groups
    p_cancel = mean(is.na(dep_time)),
    #average departure delay per groups
    dd_ave   = mean(dep_delay, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  ggplot(aes(x = dd_ave, y = p_cancel)) +
  geom_point()

```

## 3.3. SUMMARIES

a\) What will sum(is.na(x)) tell you? How about mean(is.na(x))?
```{r 3_3_1_logical}
set.seed(123)
v_length <- 9
v <- 1:v_length
NA_index <- sample(1:v_length, size = (1/3)*v_length, replace = FALSE)
v[NA_index] <- NA
#v <- (1, 2, NA, 4, 5, NA, 7, 8, NA)
v
sum(is.na(v))
mean(is.na(v))

```

b\) What does prod() return when applied to a logical vector? What logical summary function is it equivalent to? What does min() return when applied to a logical vector? What logical summary function is it equivalent to? Read the documentation and perform a few experiments.
```{r 3_3_2_logical, eval=FALSE}
?prod
?max

```
```{r 3_3_3_logical}
df <- tibble(
  v1 = c(TRUE,  TRUE,  TRUE),
  v2 = c(FALSE, FALSE, FALSE),
  v3 = c(TRUE,  FALSE, FALSE)
)
df

#prod (all)
df |> summarize(prod_v1 = prod(v1), prod_v2 = prod(v2), prod_v3 = prod(v3))

#max (any)
df |> summarize(max_v1 = max(v1), max_v2 = max(v2), max_v3 = max(v3))

```

## 3.4. CONDITIONAL TRANSFORMATIONS

a\) A number is even if it’s divisible by two, which in R you can find out with x %% 2 == 0. Use this fact and if_else() to determine whether each number between 0 and 20 is even or odd.
```{r 3_4_1_logical}
tibble(numbers = 1:20) |>
  mutate(eve_odd = if_else(numbers %% 2 == 0, "even", "odd"))

```

b\) Given a vector of days like x <- c("Monday", "Saturday", "Wednesday"), use an if_else() statement to label them as weekends or weekdays.
```{r 3_4_2_logical}
set.seed(123)
wdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday",
  "Sunday")
tibble(wday = sample(wdays, 15, rep=TRUE)) |>
  mutate(wlab = if_else(
    wday %in% c("Saturday", "Sunday"),
    "weekends",
    "weekdays")
  )

```

c\) Use if_else() to compute the absolute value of a numeric vector called x.
```{r 3_4_3_logical}
v <- -6:6
if_else(v < 0, -v, v)

```

d\) Write a case_when() statement that uses the month and day columns from flights to label a selection of important US holidays (e.g., New Years Day, 4th of July, Thanksgiving, and Christmas). First create a logical column that is either TRUE or FALSE, and then create a character column that either gives the name of the holiday or is NA.
```{r 3_4_4_logical}
flights |>
  mutate(
    is_important =
      month ==  1 & day ==  1 |        #New Years Day
      month ==  4 & day ==  7 |        #4th of July
      month == 11 & day == 28 |        #Thanksgiving 2013
      month == 12 & day == 25,         #Christmas
    important_day = case_when(
      month ==  1 & day ==  1 ~ "New Years Day",
      month ==  4 & day ==  7 ~ "4th of July",
      month == 11 & day == 28 ~ "Thanksgiving",
      month == 12 & day == 25 ~ "Christmas"
    ),
    .keep = "used"
  ) |>
  group_by(is_important,important_day) |>
  count()

```
