---
title: "3.modelling (exercises)"
author: "Joao Lopes"
date: "2026-03-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, eval=FALSE)

```

\newpage

\tableofcontents

***
\newpage

# 1. BUILD A MODEL

```{r 1_variation, warning=FALSE, message=FALSE}
library("dotwhisker") 
library("tidymodels")
library("tidyverse")
library("janitor")

```

## 1.1. LOOK AT DATA

[no exercises]

## 1.2. FIT A MODEL

a\) Evaluate the following model using function glance(). Evaluate it also using function augment() and rsq() from tidymodels. What are the possible problems of using the same data to fit and to evaluate the model?
```{r 1_2_1_modelling, echo=TRUE, eval=FALSE}
set.seed(123)
lev_cut = c("Fair", "Good", "Very Good", "Premium", "Ideal")
diamonds_data <- diamonds |>
  mutate(
    log_price = log(price),
    log_carat = log(carat),
    cut = fct(as.character(cut), levels = lev_cut),
    .keep = "used"
  ) |>
  slice_sample(n = 1000)
lm_fit <- linear_reg() |>
  fit(log_price ~ log_carat + cut, data = diamonds_data)

```
```{r 1_2_2_modelling}
#loading data and making changes
set.seed(123)
lev_cut = c("Fair", "Good", "Very Good", "Premium", "Ideal")
diamonds_data <- diamonds |>
  mutate(
    log_price = log(price),
    log_carat = log(carat),
    cut = fct(as.character(cut), levels = lev_cut),
    .keep = "used"
  ) |>
  slice_sample(n = 1000)

#fit model
lm_fit <- linear_reg() |>
  fit(log_price ~ log_carat + cut, data = diamonds_data)

#evaluate using glance()
glance(lm_fit)

#evaluate using augment() and rsq()
augment(lm_fit, new_data = diamonds_data) |>
  rsq(truth = log_price, estimate = .pred)

# Conclusions
# 1. Using same data to fit and evaluate can overestimate evaluation.

```

b\) How can you avoid overestimating the evaluation of the model? Consider taking a new sample of the data.
```{r 1_2_3_modelling}
#obtain a new sample of the data
set.seed(456)
diamonds_data2 <- diamonds |>
  mutate(
    log_price = log(price),
    log_carat = log(carat),
    cut = fct(as.character(cut), levels = lev_cut),
    .keep = "used"
  ) |>
  slice_sample(n = 1000)

#evaluate using the new sample
augment(lm_fit, new_data = diamonds_data2) |>
  rsq(truth = log_price, estimate = .pred)

```

c\) Using the same fitted model, build by hand a plot showing how the residuals vary a long the values of the predictors. 
```{r 1_2_4_modelling}
#plot residuals ~ fitted
augment(lm_fit, new_data = diamonds_data) |>
  ggplot(aes(y = .resid, x = .pred)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue")

```

## 1.3. USE MODEL TO PREDICT

a\) Using the model from the previous section, predict the price of diamonds with 1, 2 and 3 carat for cuts "Fair", "Very Good" and "Ideal". Build a plot of the predictions using a confidence interval of 90%.
```{r 1_3_1_modelling}
#create new points
new_points2 <- expand.grid(
  log_carat = log(1:3),  cut = c("Fair", "Very Good", "Ideal"))

#create plot
bind_cols(
  #new points
  new_points2,
  #fitted values
  predict(lm_fit, new_data = new_points2),
  #confidence intervals
  predict(lm_fit, new_data = new_points2, type = "conf_int", level = 0.90)
) |>
  #plot data
  ggplot(aes(x = exp(log_carat), color = cut)) + 
  geom_point(aes(y = exp(.pred))) + 
  geom_errorbar(
    aes(ymin = exp(.pred_lower), ymax = exp(.pred_upper)),
    width = 0.1
  ) + 
  labs(x = "diamond carat", y = "diamond price", color = "diamond cut")

```

***
\newpage

# 2. BUILD A MODEL USING WORKFLOW

```{r 2_modelling, warning=FALSE, message=FALSE}
library("nycflights13")
library("tidymodels")
library("tidyverse")

```

## 2.1. LOOK AT DATA

[no exercises]

## 2.2. SPLIT DATA

[no exercises]

## 2.3. CREATE MODEL

[no exercises]

## 2.4. CREATE RECIPE

[no exercises]

## 2.5. CREATE WORKFLOW

a\) Considering the following workflow, build a new workflow without the steps of the recipe. Consider replacing the function add_recipe() with add_formula().
```{r 2_3_1_modelling, echo=TRUE, eval=FALSE}
glm_mod <- logistic_reg() |>
  set_engine("glm")
glm_rec <- 
  recipe(arr_delay ~ ., data = train_data) |>
  step_date(date, features = c("dow", "month")) |>
  step_holiday(date, holidays = timeDate::listHolidays("US")) |>
  step_rm(date) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())
glm_wflow <-
  workflow() |>
  add_model(glm_mod) |>
  add_recipe(glm_rec)

```
```{r 2_3_2_modelling}
#create model specification
glm_mod <- logistic_reg() |>
  set_engine("glm")

#create workflow
glm_wflow2 <-
  workflow() |>
  add_model(glm_mod) |>
  add_formula(arr_delay ~ .)

```

## 2.6. FIT A MODEL

a\) Using the given workflow and the newly created one, fit the two models to the following data set. Consider the data split in train and test sets.
```{r 2_4_1_modelling, echo=TRUE, eval=FALSE}
set.seed(123)
flight_data <-
  flights |> 
  mutate(
    arr_delay = if_else(arr_delay >= 30, "late", "on_time"),
    arr_delay = fct(arr_delay, levels = c("late", "on_time")),
    date = lubridate::as_date(time_hour)
  ) |> 
  select(dep_time, origin, dest, distance, carrier, date, arr_delay) |> 
  drop_na() |> 
  mutate(across(where(is.character), as.factor)) |>
  slice_sample(n = 10000)

set.seed(456)
data_split <- initial_split(flight_data, prop = 3/4)
train_data <- training(data_split)
test_data  <- testing(data_split)

```
```{r 2_4_2_modelling}
#loading data and making changes
set.seed(123)
flight_data <- 
  flights |> 
  mutate(
    arr_delay = if_else(arr_delay >= 30, "late", "on_time"),
    arr_delay = fct(arr_delay, levels = c("late", "on_time")),
    date = lubridate::as_date(time_hour)
  ) |> 
  select(dep_time, origin, dest, distance, carrier, date, arr_delay) |> 
  drop_na() |> 
  mutate(across(where(is.character), as.factor)) |>
  slice_sample(n = 10000)

#create random data split
set.seed(456)
data_split <- initial_split(flight_data, prop = 3/4)
train_data <- training(data_split)
test_data  <- testing(data_split)

#create model specification
glm_mod <- logistic_reg() |>
  set_engine("glm")

#create recipe
glm_rec <- 
  #new recipe
  recipe(arr_delay ~ ., data = train_data) |>
  #add date features
  step_date(date, features = c("dow", "month")) |>
  step_holiday(date, holidays = timeDate::listHolidays("US")) |>
  step_rm(date) |>
  #convert categorical variables to dummies
  step_dummy(all_nominal_predictors()) |>
  #remove columns with zero variance
  step_zv(all_predictors())

#create workflow 1
glm_wflow <-
  workflow() |>
  add_model(glm_mod) |>
  add_recipe(glm_rec)

#create workflow 2
glm_wflow2 <-
  workflow() |>
  add_model(glm_mod) |>
  add_formula(arr_delay ~ .)

#fit both models
glm_fit <- glm_wflow |> fit(data = train_data)
glm_fit2 <- glm_wflow2 |>  fit(data = train_data)
```

## 2.7. EVALUATE MODEL

a\) How did the previously fitted models perform? Use the appropriate statistic. 
```{r 2_5_1_modelling}
#evaluate both models
glm_aug <- augment(glm_fit, new_data = test_data)
glm_aug |> roc_auc(truth = arr_delay, .pred_late)
glm_aug2 <- augment(glm_fit2, new_data = test_data)
glm_aug2 |> roc_auc(truth = arr_delay, .pred_late)

```
b\) Looking at the augmented data set, try to guess which probability threshold was used for predicting the classes in .pred_class? 
```{r 2_5_2_modelling}
glm_aug |>
  select(.pred_class, .pred_late) |>
  arrange(desc(.pred_late)) |>
  mutate(
    is_ontime = as.numeric(.pred_class == "on_time"),
    is_thresh = as.logical(is_ontime - lag(is_ontime))
  ) |>
  filter(is_thresh)

# Conclusions:
# 1. The probability considered is the classic 50% threshold.

```

c\) Using function conf_mat(), calculate a confusion matrix for the best model. Using function tabyl(), calculate by hand a confusion matrix to the best model.
```{r 2_5_3_modelling}
#calculate confusion matrix with function conf_mat()
glm_aug |>
  conf_mat(truth = arr_delay, estimate = .pred_class)

#calculate confusion matrix by hand
glm_aug |> tabyl(.pred_class, arr_delay) |>
  rename(Prediction = .pred_class)

```

d\) Use probability thresholds of 0.40 and 0.60 to predict the classes of the best model. How does the threshold value impact on the confusion matrix? 
```{r 2_5_4_modelling}
#decrease probability threshold to 0.40
glm_aug |>
  mutate(.pred_class = if_else(.pred_on_time > 0.40, "on_time", "late")) |>
  tabyl(.pred_class, arr_delay) |>
  rename(Prediction = .pred_class)

# Conclusions 1
# 1. A lower threshold decreases true positives and increases true negatives.

#increase probability threshold to 0.60
glm_aug |>
  mutate(.pred_class = if_else(.pred_on_time > 0.60, "on_time", "late")) |>
  tabyl(.pred_class, arr_delay) |>
  rename(Prediction = .pred_class)

# Conclusions 2:
# 2. A higher threshold increases true positives and decreases true negatives.

```

e\) Using the original confusion matrix calculate the sensitivity (= TP/P) and the specificity (= TN/N). 
```{r 2_5_5_modelling}
#calculate statistics of the confusion matrix
glm_aug |> conf_mat(truth = arr_delay, .pred_class) |> summary()

#calculate sensitivity (= TP/P)
32/(32 + 338)

#calculate specificity (= TN/N)
2099/(31 + 2099)

```

f\) Create the ROC curve and place the calculated values of sensitivity and specificity on it. Use geom_hline(), geom_vline() and geom_point(). 
```{r 2_5_6_modelling}
#create ROC curve with sensitivity and specificity of the model
glm_aug |>
  roc_curve(truth = arr_delay, .pred_late) |> 
  autoplot() +
  geom_hline(yintercept = 0.086, lty = 2) +
  geom_vline(xintercept = 1 - 0.985, lty = 2) +
  geom_point(aes(x = 1 - 0.985, y = 0.086), shape=21, size=2)

```

g\) Find which terms are more important to the best fitted model and plot them. Use the statistics of each term as a proxy for their importance. Try also using the estimates of the coefficients. Are the results similar? Note that term "(Intercept)" is meaningless and should be discarded. 
```{r 2_5_7_modelling}

#using importance = statistics
glm_fit |> 
  extract_fit_parsnip() |> 
  tidy() |>
  filter(!term %in% c("(Intercept)")) |>
  mutate(
    importance = abs(statistic),
    term = fct_reorder(term, importance)
  ) |>
  arrange(desc(importance)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = importance, y = term)) +
  geom_col()

# Conclusions 1:
# 1. dep_time is by far the most important term

glm_fit |> 
  extract_fit_parsnip() |> 
  tidy() |>
  filter(!term %in% c("(Intercept)")) |>
  mutate(
    importance = estimate,
    term = fct_reorder(term, importance)
  ) |>
  arrange(desc(importance)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = importance, y = term)) +
  geom_col()

# Conclusions 2:
# 1. destinations are important terms
# 2. specific US holidays are also important

# Conclusions 3:
# 1. results are very different,
# 2. 'statistics' focus on the significance of the terms,
# 3. 'estimate' focus on the absolute value of the term.

```

***
\newpage

# 3. BUILD VARIOUS MODELS USING WORKFLOW

```{r 3_modelling, warning=FALSE, message=FALSE}
library("nnet")
library("NeuralNetTools")
library("vip")
library("tidymodels")
library("tidyverse")

```

## 3.1. LOOK AT DATA

[no exercises]

## 3.2. SPLIT DATA

[no exercises]

## 3.3. CREATE WORKFLOW AND FIT MODEL 1

a\) Considering the following split data set and tidymodel recipe, create the respective workflow and fit a simple logistic model. Use the ROC curve and ROC area to evaluate your model.
```{r 3_3_1_modelling, echo=TRUE, eval=FALSE}
hotels <- 
  read_csv("https://tidymodels.org/start/case-study/hotels.csv") |>
  mutate(across(where(is.character), as.factor))

set.seed(123)
splits      <- initial_split(hotels, strata = children, prop = 0.75)
hotel_other <- training(splits)
hotel_test  <- testing(splits)

holidays <- c("AllSouls", "AshWednesday", "ChristmasEve", "Easter", 
              "ChristmasDay", "GoodFriday", "NewYearsDay", "PalmSunday")
lr_recipe <- 
  recipe(children ~ ., data = hotel_other) |> 
  step_date(arrival_date) |> 
  step_holiday(arrival_date, holidays = holidays) |> 
  step_rm(arrival_date) |> 
  step_dummy(all_nominal_predictors()) |> 
  step_zv(all_predictors()) |> 
  step_normalize(all_predictors())

```
```{r 3_3_2_modelling}
#loading data and making changes
hotels <- 
  read_csv("https://tidymodels.org/start/case-study/hotels.csv") |>
  mutate(across(where(is.character), as.factor))

#create random data split
set.seed(123)
splits      <- initial_split(hotels, strata = children, prop = 0.75)
hotel_other <- training(splits)
hotel_test  <- testing(splits)

#create recipe
holidays <- c("AllSouls", "AshWednesday", "ChristmasEve", "Easter", 
              "ChristmasDay", "GoodFriday", "NewYearsDay", "PalmSunday")
lr_recipe <- 
  #new recipe
  recipe(children ~ ., data = hotel_other) |> 
  #add date features
  step_date(arrival_date) |> 
  step_holiday(arrival_date, holidays = holidays) |> 
  step_rm(arrival_date) |> 
  #convert categorical variables to dummies
  step_dummy(all_nominal_predictors()) |> 
  #remove columns with zero variance
  step_zv(all_predictors()) |> 
  #normalize numeric data
  step_normalize(all_predictors())

#create model specification
glm_mod <- 
  logistic_reg() |> 
  set_engine("glm")

#create workflow  
glm_workflow <- 
  workflow() |> 
  add_model(glm_mod) |> 
  add_recipe(lr_recipe)

#fit model using training data
t1 <- Sys.time()
glm_res <- 
  glm_workflow |> 
  fit(data = hotel_other)
Sys.time() - t1

#calculate predictions using testing data
glm_aug <- augment(glm_res, new_data = hotel_test)

#evaluate model using ROC curve
glm_aug |> roc_curve(children, .pred_children) |> autoplot()

#evaluate model using ROC area
glm_aug |> roc_auc(truth = children, .pred_children)

```

## 3.4. CREATE WORKFLOW AND FIT MODEL 2

a\) Replace the glm model used above with a neural network model. Consider function mlp() with engine "nnet" and mode "classification". For the hyper parameters, set hidden_units = 1 and penalty = tune(). Use the same recipe as before. To tune the model set a grid of 10 points and use the given data split. Choose the best model and evaluate it with a ROC curve and the ROC area. How confident are you about the tuning?
```{r 3_4_1_modelling, echo=TRUE, eval=FALSE}
set.seed(234)
val_set <- validation_split(hotel_other, strata = children, prop = 0.80)
```
```{r 3_4_2_modelling}
#create random split of training data
set.seed(234)
val_set <- validation_split(hotel_other, strata = children, prop = 0.80)

#create model specification
cores <- parallel::detectCores()
nn_mod <- 
  mlp(
    hidden_units = 1, #hidden units of the neural network
    penalty = tune(), #amount of regularization
  ) |> 
  set_engine("nnet", num.threads = cores) |> 
  set_mode("classification")

#create workflow
nn_workflow <- 
  workflow() |> 
  add_model(nn_mod) |> 
  add_recipe(lr_recipe)

#fit model using training data and evaluate using tuning data
t1 <- Sys.time()
set.seed(345)
nn_res <- 
  nn_workflow |> 
  tune_grid(
    val_set,
    grid = 10,
    control = control_grid(save_pred = TRUE),
    metrics = metric_set(roc_auc)
  )
Sys.time() - t1

#plot tuning metrics
autoplot(nn_res)

# Conclusions 1
# 1. the best values of the parameter seem to be rather random,
# 2. results provide little confidence on the tuning.

#select best tuning parameter
nn_best <- 
  nn_res |> 
  select_best(metric = "roc_auc")

#collect predictions using best tuned model
nn_pred <-
  nn_res |> 
  collect_predictions(parameters = nn_best)

#evaluate model using ROC curve
nn_pred |> 
  roc_curve(children, .pred_children) |> 
  autoplot()

#evaluate model using ROC area
nn_pred |> roc_auc(children, .pred_children)

```

## 3.5. EVALUATE MODEL

a\) Choose the best tuned neural network model and evaluated it using the test data created before. For the evaluation, consider the ROC curve and ROC area. Finish the analysis by plotting the most important features (i.e. predictors) of the final model. Set the importance parameter of the engine to "impurity". 
```{r 3_5_1_modelling}
#build last model
last_nn_mod <- 
  mlp(hidden_units = 1, penalty = nn_best$penalty) |> 
  set_engine("nnet", num.threads = cores, importance = "impurity") |> 
  set_mode("classification")

#create last workflow
last_nn_workflow <- 
  #neural network workflow
  nn_workflow |> 
  #update with tuned model
  update_model(last_nn_mod)

#fit last model using testing data
t1 <- Sys.time()
set.seed(345)
last_nn_fit <- 
  last_nn_workflow |> 
  last_fit(split = splits)
Sys.time() - t1

#collect predictions using testing data
last_rf_pred <- last_nn_fit |>
  collect_predictions()

#evaluate model using ROC curve
last_rf_pred |>
  roc_curve(children, .pred_children) |>
  autoplot()

#evaluate model using ROC area
last_rf_pred |>
  roc_auc(children, .pred_children)

#extract results
last_nn_fit |> 
  extract_fit_parsnip() |> 
  vip(num_features = 20)

```
